---
title: "Data Monitoring Tutorial"
project:
  type: default
  output-dir: docs
format:
  gfm:
    toc: true
date: 2026-06-11
author: Allie Warren, allison.warren
editor: visual
prefer-html: true
---

## Overview

This tutorial shows a basic monitoring workflow for weekly reported public health count data.

The example synthetic data are cumulative delayed weekly counts:

-   `report_week`: week when the data extract/report was produced

-   `reference_week`: week when the events occurred

-   `ach_region`: Washington Accountable Community of Health region

-   `sex`: `Female` or `Male`

    `counts`: cumulative count for that `reference_week`, `ach_region`, and `sex` as of the `report_week`.

The data has been synthetically produced with some errors and assumes counts are reported with some lag time (and a max lag of 10 weeks).

The tutorial includes:

1.  Rule-based validation with `validate`
2.  One category-distribution check using a chi-squared test
3.  Count monitoring to check that counts are increasing, fall within historical quantiles, and for sustained drift detection
4.  Reporting-delay monitoring using percent of final counts

```{r}
# use pacman for loading/installing the other packages
if(!require("pacman")) install.packages("pacman") 

# This load packages
pacman::p_load(readr, dplyr, tidyr, stringr, ggplot2, lubridate, validate, datadriftR)

set.seed(123)
```

```{r}

current_report_week <- ymd("2026-06-08")
# set max data delay to 10 weeks
max_delay_weeks <- 10L
# get most recent date for complete data
latest_complete_reference_week <- current_report_week - ((max_delay_weeks-1)*7)
# get the previous 12 weeks of complete data - which we will evaluate compared to historic data
recent_completed_start <- latest_complete_reference_week - (12*7)
  
# expected categories for ACH regions and sex fields
expected_ach_regions <- c(
  "Olympic Community of Health",
  "North Sound ACH",
  "North Central ACH",
  "Better Health Together",
  "HealthierHere",
  "Elevate Health",
  "Cascade Pacific Action Alliance",
  "SWACH",
  "Greater Columbia ACH"
)

expected_sex_values <- c("Female", "Male")
```

## Read in Data

```{r}
synthetic_counts <- read_csv("../data/synthetic_reporting_data.csv")

current_extract <- synthetic_counts |>
  filter(report_week == current_report_week)

head(current_extract)
```

```{r}
# create format for data alert data
alert_table <- tibble(
    check_type = character(),
    alert_level = character(),
    message = character())
```

## Validate expected values and categories

The `validate` package is useful when you want explicit rules that can be run repeatedly on each incoming data extract.

Here we check:

-   `reference_week` is not after `report_week`;
-   `counts` is present and nonnegative;
-   `sex` is either `Female` or `Male` and not NA
-   `ach_region` is one of the expected ACH regions.

```{r}
# create rule set
validation_rules <- validator(
  reference_week <= report_week,
  !is.na(counts),
  counts >= 0,
  !is.na(sex),
  sex %in% expected_sex_values,
  !is.na(ach_region),
  ach_region %in% expected_ach_regions)

# apply ruleset to the dataset
validation_results <- confront(synthetic_counts, validation_rules)

# view if there are errors in the data
validation_summary <- summary(validation_results)
validation_summary |> select(items, fails, expression)

# add error to alert table
validation_alerts <- validation_summary |>
  filter(fails > 0) |>
  # transmute creates a new data frame with only the newly created fields
  transmute(check_type = "validation checks",
            alert_level = "warning",
            message = str_glue("Data validation error: {expression}"))
alert_table <- rbind(alert_table, validation_alerts)

```

view the violating data rows

```{r}
error_rows <- violating(synthetic_counts, validation_results)
error_rows

# Get logical matrix of passes/fails (rows = observations, cols = rules)
status_matrix <- values(validation_results)

# Row indices that fail ANY rule
error_indices <- which(rowSums(!status_matrix, na.rm = TRUE) > 0)

# remove the error rows
synthetic_counts_clean <- synthetic_counts[-error_indices,]

```

## Check whether the distribution of a category changed

For this tutorial, we compare the sex distribution in the most recent completed 12 reference weeks to the historical baseline.

Because counts are delayed, we use only reference weeks that are at least `max_delay_weeks` old. That means delay week `max_delay_weeks` is observed and can be treated as final.

The chi-squared test asks whether the distribution of counts across `Female` and `Male` differs between the historical and recent completed periods.

```{r}
# filter to complete data
finalized_counts <- synthetic_counts_clean |>
  # removing the most recent 9 weeks of data, which are incomplete
  filter(reference_week <= latest_complete_reference_week,
         # filter to final counts - cases reported as of 10 weeks after ref data
         report_week == reference_week + weeks(max_delay_weeks))

# get proportion of counts for each sex for historic reporting period and more recent period
sex_distribution <- finalized_counts |>
  mutate(period = if_else(reference_week >= recent_completed_start,
                          "Recent completed weeks",
                          "Historical baseline")) |>
  summarise(counts = sum(counts), .by = c(period, sex)) |>
  mutate(prop = counts / sum(counts), .by = period)

sex_distribution
```

```{r}
sex_matrix <- sex_distribution |>
  select(period, sex, counts) |>
  pivot_wider(names_from = sex, values_from = counts, values_fill = 0) |>
  tibble::column_to_rownames("period") |>
  as.matrix()

# chi square test
sex_chisq <- chisq.test(sex_matrix)

sex_chisq_summary <- tibble(statistic = unname(sex_chisq$statistic),
                            df = unname(sex_chisq$parameter),
                            p_value = sex_chisq$p.value)

sex_chisq_summary

# add to alerts table if the distribution likely changed
sex_distribution_alerts <- sex_chisq_summary |>
  filter(p_value < 0.01) |>
  transmute(check_type = "category_distribution",
            alert_level = "warning",
            message = str_glue("Chi-squared test suggests the recent sex distribution changed: p = {signif(p_value, 3)}"))

alert_table <- rbind(alert_table, sex_distribution_alerts)
```

visualize the change in distribution

```{r}
sex_distribution |>
  ggplot(aes(x = period, y = prop, fill = sex)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  labs(title = "Sex distribution: historical baseline vs recent completed weeks",
       x = NULL,
       y = "Percent of counts",
       fill = "Sex")
```

## Cumulative counts should not decrease

For a fixed `reference_week`, `ach_region`, and `sex`, cumulative counts should only increase or stay the same as `report_week` advances (this assumption is likely not true in real data, but it is good to identify in your data as it may cause issues).

```{r}
# check whether the current weekly counts (for a given region and group) have decreased compared to the previous week
decreasing_counts_issues <- synthetic_counts |>
  filter(sex %in% expected_sex_values,
         ach_region %in% expected_ach_regions) |>
  arrange(ach_region, sex, reference_week, report_week) |>
  mutate(previous_counts = lag(counts),
         decrease = counts < previous_counts,
         .by = c(ach_region, sex, reference_week)) |>
  filter(decrease)

decreasing_counts_issues

# add to alert table
decreasing_counts_alerts <- decreasing_counts_issues |>
  transmute(check_type = "reporting_delay",
            alert_level = "critical",
            message = str_glue("Cumulative counts decreased from {previous_counts} to {counts} as reporting advanced to {report_week}"))

alert_table <- rbind(alert_table, decreasing_counts_alerts)

```

visualize examples of the decreasing counts

```{r}

if (nrow(decreasing_counts_issues) > 0) {

  decreasing_counts_data <- inner_join(synthetic_counts_clean, decreasing_counts_issues |> select(-report_week), by = c('reference_week', 'ach_region', 'sex'))

  decreasing_counts_data |>
    ggplot(aes(x = report_week, y = counts.x, color = paste0(ach_region, ", ", sex))) +
    geom_line() +
    geom_point() +
    theme_bw() +
    labs(title = "Example cumulative-count decrease",
         x = "Report week",
         y = "Cumulative counts", color = 'Data Grouping') +
    facet_wrap(vars(ach_region), scales = "free_x")

}
```

## Monitor counts against historical ranges

Next we aggregate finalized counts by ACH region and reference week. We compare recent completed values to each ACH region's historical 5th and 95th percentile range.

This is a simple range-based monitoring rule:

-   below historical 5th percentile: unusually low;
-   above historical 95th percentile: unusually high.

```{r}
# get weekly counts per ach region - using counts reported by 10 weeks after the reference data
ach_week_counts <- finalized_counts |>
  summarise(counts = sum(counts), .by = c(reference_week, ach_region))

count_bounds <- ach_week_counts |>
  filter(reference_week < recent_completed_start) |>
  summarise(median_count = median(counts),
            lower_bound = quantile(counts, 0.05),
            upper_bound = quantile(counts, 0.95), .by = ach_region)

# compare historic count ranges to the current counts
recent_count_monitoring <- ach_week_counts |>
  filter(reference_week >= recent_completed_start) |>
  left_join(count_bounds, by = join_by(ach_region)) |>
  mutate(count_alert = counts < lower_bound | counts > upper_bound,
         direction = case_when(counts < lower_bound ~ "low",
                               counts > upper_bound ~ "high",
                               TRUE ~ "normal"))

# view coutns outside of the normal range
recent_count_monitoring |>
  filter(count_alert)

# add to alert table
count_alerts <- recent_count_monitoring |>
  filter(count_alert) |>
  transmute(check_type = "count_range",
            alert_level = "warning",
            message = str_glue("{ach_region} count was {direction}: {counts} vs expected range ",
      "{round(lower_bound)}-{round(upper_bound)} for week {reference_week}"))

alert_table <- rbind(alert_table, count_alerts)

```

visualize counts compared to historic ranges

```{r}

ach_week_counts |> left_join(count_bounds, by = join_by(ach_region)) |>
  ggplot(aes(x = reference_week, y = counts)) +
  geom_ribbon(aes(ymin = lower_bound, ymax = upper_bound), alpha = 0.15) +
  geom_line() +
  # add vertical line for the mark the recent counts that we are evaluating
  geom_vline(aes(xintercept = recent_completed_start), color = 'blue', linetype = 'dashed') +
  # add points for counts that fall outside of the bounds
  geom_point(data = recent_count_monitoring |> filter(count_alert),
             aes(x = reference_week, y = counts), color = 'red') +
  theme_bw() +
  # create separate plots for each region
  facet_wrap(vars(ach_region), scales = "free_y") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = "ACH weekly counts with historical 5th–95th percentile ranges",
       subtitle = "Points mark recent completed weeks outside historical ACH-specific ranges",
       x = NULL,
       y = "Finalized weekly counts")

```

## Detect sustained drift with Page-Hinkley

Quantile checks are good at finding individual unusual values. Page-Hinkley is more useful for detecting sustained shifts in a numeric stream. It continuously monitors the cumulative deviation of incoming data from its running mean. If the deviation exceeds a predefined threshold it signals a change in the data distribution.

Here we apply `datadriftR::detect_drift()` to statewide finalized weekly counts. To make tuning less dependent on the absolute count scale we z-score the counts.

```{r}
statewide_counts <- ach_week_counts |>
  summarise(counts = sum(counts), .by = reference_week)

hist_stats <- statewide_counts |>
  filter(reference_week < recent_completed_start) |>
  summarise(mu = mean(counts), sigma = sd(counts))

statewide_counts <- statewide_counts |>
  mutate(count_z = (counts - hist_stats$mu) / hist_stats$sigma)

page_hinkley_results <- datadriftR::detect_drift(stream = statewide_counts$count_z,
                                                 method = "page_hinkley",
                                                 delta = 0.01,    # allowable slack in SD units
                                                 threshold = 2.0,     # ~2 SD cumulative shift triggers alarm
                                                 min_instances = 15) |>
  as_tibble()

page_hinkley_points <- page_hinkley_results |>
  filter(type == "drift") |>
  mutate(reference_week = statewide_counts$reference_week[index],
         count_index = statewide_counts$count_index[index])

page_hinkley_points

# add to alert table
page_hinkley_alerts <- page_hinkley_points |>
  transmute(check_type = "page_hinkley",
            alert_level = "info",
            message = str_glue("Page-Hinkley detected sustained drift around {reference_week} in statewide counts"))

alert_table <- rbind(alert_table, page_hinkley_alerts)

```

-   Page-Hinkley flags points where the stream appears to have shifted persistently.
-   It is sensitive to tuning choices such as `delta`, `threshold`, and `min_instances`.
-   It is usually better to run it on a monitored metric, rather than raw counts with very different scales.

Visualize:

```{r}
statewide_counts |>
  ggplot(aes(x = reference_week, y = count_z)) +
  geom_line() +
  geom_point(data = page_hinkley_points,
             aes(x = reference_week, y = value), color = 'red') +
  theme_bw() +
  labs(title = "Page-Hinkley drift detection on z-scored statewide counts",
       x = NULL,
       y = "Statewide z-scored counts") 
```

## Monitor reporting delay using percent of final counts

For cumulative delayed data, counts for a reference week should increase as more report weeks arrive.

For completed reference weeks, define:

\[ \text{percent final at delay } d = \frac{\text{cumulative count at delay } d}{\text{cumulative count at delay } 10} \]

This uses only reference weeks that are at least 10 weeks old, so the delay-10 value is observed.

```{r}

delay_percent <- synthetic_counts_clean |>
  filter(reference_week <= latest_complete_reference_week,
         sex %in% expected_sex_values,
         ach_region %in% expected_ach_regions) |>
  mutate(delay_week = as.integer(report_week - reference_week) / 7L) |>
  filter(delay_week >= 0, delay_week <= max_delay_weeks) |>
  summarise(counts = sum(counts), .by = c(reference_week, delay_week)) |>
  left_join(finalized_counts |>
              summarise(final_counts = sum(counts), .by = reference_week),
            by = join_by(reference_week)) |>
  filter(final_counts > 0) |>
  mutate(pct_final = counts / final_counts,
         period = if_else(reference_week >= recent_completed_start,
                          "Recent completed weeks",
                          "Historical baseline"))

delay_percent |>
  arrange(desc(reference_week), delay_week) |>
  head(15)


```

compare current data delays compared to historic delays

```{r}
historical_delay_band <- delay_percent |>
  filter(period == "Historical baseline") |>
  summarise(historical_mean = mean(pct_final),
            historical_p05 = quantile(pct_final, 0.05),
            historical_p95 = quantile(pct_final, 0.95),
            .by = delay_week)

recent_delay_line <- delay_percent |>
  filter(period == "Recent completed weeks") |>
  summarise(recent_mean = mean(pct_final), .by = delay_week)

delay_monitoring <- recent_delay_line |>
  left_join(historical_delay_band, by = join_by(delay_week)) |>
  mutate(slow_reporting_alert = recent_mean < historical_p05)

delay_monitoring

# add to alert table
delay_alerts <- delay_monitoring |>
  filter(slow_reporting_alert) |>
  transmute(check_type = "reporting_delay",
            alert_level = "warning",
            message = str_glue("Recent percent final is low at delay week {delay_week}: ",
                               "{scales::percent(recent_mean, accuracy = 0.1)} vs historical 5th percentile ",
                               "{scales::percent(historical_p05, accuracy = 0.1)}"))

alert_table <- rbind(alert_table, delay_alerts)

```

visualize:

```{r}
historical_delay_band |>
  ggplot(aes(x = delay_week)) +
  geom_ribbon(aes(ymin = historical_p05, ymax = historical_p95), alpha = 0.15) +
  geom_line(aes(y = historical_mean), linetype = "dashed") +
  geom_line(data = recent_delay_line,
            aes(y = recent_mean)) +
  geom_point(data = delay_monitoring |> filter(slow_reporting_alert),
             aes(y = recent_mean)) +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  labs(title = "Reporting delay: percent of final counts by delay week",
       subtitle = "Ribbon = historical 5th–95th percentile; solid line = recent completed weeks",
       x = "Delay week",
       y = "Percent of final count" )
```

## View All Alerts

```{r}
alert_table
```
